---
output:
  md_document:
    variant: markdown_github
---

# Purpose

Purpose of this work folder is to analyse South African export data and the relationship between exports, foreign reserves and the Exchange rate .




```{r}

rm(list = ls()) # Clean your environment:
gc() # garbage collection - It can be useful to call gc after a large object has been removed, as this may prompt R to return memory to the operating system.
library(tidyverse)
list.files('code/', full.names = T, recursive = T) %>% .[grepl('.R', .)] %>% as.list() %>% walk(~source(.))
```

```{r}
# Loading libraries
pacman::p_load(tidyverse, RColorBrewer,zoo,imfapi)
```

# 1. Data Importing
## 1.1 SA exports
```{r}
sa_exports <- imf_get(
    dataflow_id = "IMTS",
    dimensions = list(
        FREQUENCY = "M",             # Monthly
        COUNTRY = "ZAF",             # South Africa
        COUNTERPART_COUNTRY = "",    # "" is World (Total Exports per country), while "G001" is aggregate
        INDICATOR = "XG_FOB_USD"     # Exports, Goods, US Dollars
    ),
    start_period = "2010-Q1"       # No end period, that the data is pulled up till most recent period
)

head(sa_exports)
```

## 1.2 SA foreign reserves
```{r}
sa_foreign_reserves <- imf_get(
    dataflow_id = "IL",
    dimensions = list(
        COUNTRY = "ZAF",              # South Africa
        INDICATOR = c("RXF11FX_REVS", # Reserves excluding gold, foreign exchange
                      "NFAOFA_ACO_NRES_S121", # Net foreign assets, Claims on Nonresidents, Other depository corporations
                      "NFAOFL_LT_NRES_S121",  # Net foreign assets, other foreign liabilities, Liabilities to Nonresidents, Central bank
                      "TRGMV_REVS") ,
        UNIT = "USD",                # Measurement currency
        FREQUENCY = "M"             # Monthly
    ),
    start_period = "2010-Q1"       # No end period, that the data is pulled up till most recent period
)

head(sa_foreign_reserves)
```

## 1.3 EFFECTIVE EXCAHNGE RATE DATA
```{r}
sa_EER <- imf_get(
    dataflow_id = "EER",
    dimensions = list(
        COUNTRY = "ZAF",                         # South Africa
        INDICATOR = c("REER_IX_RY2010_ACW_RCPI", # Real
                      "NEER_IX_RY2010_ACW"),     # Nominal
        FREQUENCY = "M"                          # Monthly
    ),
    start_period = "2010-Q1"                     # No end period, that the data is pulled up till most recent period
)

head(sa_EER)
```

## 1.4 SA_EUR_USD exchange rate data
```{r}
sa_FX <- imf_get(
    dataflow_id = "ER",
    dimensions = list(
        COUNTRY = "ZAF",                         # South Africa
        INDICATOR = c("XDC_EUR",
                      "XDC_USD"),
        TYPE_OF_TRANSFORMATION = c("EOP_RT","PA_RT"),
        FREQUENCY = "M"                          # Monthly
    ),
    start_period = "2010-Q1"                     # No end period, that the data is pulled up till most recent period
)

head(sa_FX)
```

# 2 Data wrangling
## 2.1 sa_EER
```{r}
sa_EER <- sa_EER %>% select(-FREQUENCY)

```
Just removing redundant variable

## 2.2 sa_foreign_reserves
```{r}
sa_foreign_reserves <- sa_foreign_reserves %>% select(-c("FREQUENCY", "UNIT")) %>%
    spread(INDICATOR, OBS_VALUE) %>%
    rename(Reserves_excl_gold = RXF11FX_REVS) %>%
    select(COUNTRY, TIME_PERIOD, Reserves_excl_gold) %>%
    mutate(reserves_indexed = Reserves_excl_gold/32309728657) # reserves indexed to 2010
```
Chose desired variables, renamed them and indexed reserves to the starting value of the data Jan 2010

## 2.3 sa_FX
```{r}
sa_FX <- sa_FX %>%
    pivot_wider(
        names_from = c(INDICATOR, TYPE_OF_TRANSFORMATION),
        values_from = OBS_VALUE,
        names_sep = "_") %>%
    rename(ZAR_EUR_PA = XDC_EUR_PA_RT,
           ZAR_EUR_EOP = XDC_EUR_EOP_RT,
           ZAR_USD_PA = XDC_USD_PA_RT,
           ZAR_USD_EOP = XDC_USD_EOP_RT)
```

Made the data wide and renamed the variables.

## 2.4 sa_exports
```{r}
# sa_exports
sa_exports <- sa_exports %>%
    # 1. Remove Frequency
    select(-FREQUENCY) %>%

    # 2. Filter out regions with numbers (using your regex)
    filter(!grepl("[0-9]", COUNTERPART_COUNTRY)) %>%

    # 3. Create the date and year columns FIRST
    mutate(date_fixed = lubridate::ym(TIME_PERIOD),
           year = lubridate::year(date_fixed)) %>%

    # 4. NOW you can safely remove the old columns
    select(-date_fixed) %>%

    # 5. Rename the indicator
    mutate(INDICATOR = "Exports_USD")



# Analysis
sa_exports_ranked <- sa_exports %>%
    # 1. Sum values for each country per year
    group_by(COUNTERPART_COUNTRY, year) %>%
    summarise(annual_exp = sum(OBS_VALUE, na.rm = TRUE), .groups = "drop") %>%

    # 2. Rank countries within each year
    group_by(year) %>%
    mutate(export_rank = rank(desc(annual_exp))) %>%
    ungroup() %>%

    # Calculating each country's percentage of total exports per year
    group_by(year) %>%
    mutate(annual_exp_perc = (annual_exp/sum(annual_exp))*100) %>%
    arrange(year, annual_exp_perc)

# aggregating exports by year and indexing
sa_exports_total <- sa_exports %>% 
    ungroup() %>%
    group_by(TIME_PERIOD) %>%
    summarise(tot_mon_exports = sum(OBS_VALUE, na.rm = T)) %>%

    # Indexed to 2010-m01
    mutate(tot_export_index = tot_mon_exports/5279270626)

```
Filtered, ranked, aggregated and indexed the data

## 2.5 Series of joins to finalise plotting data
```{r}
df_final <- calc_rolling_vol(sa_FX, TIME_PERIOD, ZAR_USD_PA,12)
 df_final <- df_final %>% slice(-(1:12))
sa_foreign_reserves <- sa_foreign_reserves %>% slice(-(1:12))


scatterplot_data <- left_join(df_final,sa_foreign_reserves, by = "TIME_PERIOD")

# 1. Filter the EER data FIRST to avoid the many-to-many explosion
sa_REER_only <- sa_EER %>%
    filter("REER_IX_RY2010_ACW_RCPI" == INDICATOR)

scatterplot_data <- scatterplot_data %>%
    left_join(sa_REER_only, by = "TIME_PERIOD") %>%
    # Use everything() first to see where the column went,
    # or use this specific select that accounts for renaming:
    select(
        TIME_PERIOD,
        rolling_vol,
        reserves_indexed,
        # This picks up INDICATOR even if R renamed it to INDICATOR.y
        any_of(c("OBS_VALUE", "OBS_VALUE.y", "OBS_VALUE.x")),
        Reserves_excl_gold
    )
scatterplot_data %>%  left_join(sa_exports_total)

```

# 3. Plots
AI did give me the explanations. I will when I have time add my own interpretation.
## Plot 1 
```{r}
# 1. Update the data 
scatterplot_data <- scatterplot_data %>%
    mutate(date_axis = ym(TIME_PERIOD))

# 2. Re-create the plot object using the NEW column 'date_axis'
g <- plot_scatter(scatterplot_data, x = date_axis, y = OBS_VALUE.y, color = reserves_indexed)

# 3. Apply the date formatting and labels
g <- g + 
    scale_x_date(date_breaks = "2 years", date_labels = "%Y") +
    labs(
        x = "Timeline", # Updated from 'Reserve Levels' because X is now a date
        y = "REER (Competitiveness)",
        title = "The Reserve Buffer vs. Real Exchange Rate"
    )

# 4. View it
g

```
The Competitiveness Trend: If the line (REER) is trending upward, South Africa is becoming less competitive. It means our goods are becoming more expensive relative to our trading partners.

The Reserve Response: Look for periods where the REER spikes and the color (Reserves) changes significantly. If the color fades (lower reserves) while REER is high, it suggests the SARB lacks the "ammunition" to intervene and weaken the Rand to help exporters.

## Plot 2
```{r}
scatterplot_data <- scatterplot_data %>%  left_join(sa_exports_total)

g2 <- plot_scatter(scatterplot_data, reserves_indexed, rolling_vol , color = tot_export_index)
g2 <- g2 + 
    labs(
        title = "Currency Volatility vs. Reserve Adequacy",
        subtitle = "Analyzing the relationship between ZAR volatility and SARB reserve levels",
        x = "Reserves (Indexed)",
        y = "12-Month Rolling Volatility (ZAR/USD)",
        color = "Export Performance",
        caption = "Source: IMF monthly data | Calculations: Log-returns rolling SD"
    ) +
    theme_minimal() # Optional: keeps it clean for academic printing

g2
```
The "Defensive Cap" Evidence: You want to see a negative slope. As you move to the right (Higher Reserves), the points should settle lower on the Y-axis (Lower Volatility). This proves that a larger "war chest" successfully caps exchange rate uncertainty.

The Danger Zone: Look at the top-left corner. If you have many points there, it shows "Low Reserves + High Volatility." This is where exporters struggle most because they can't price their products reliably.

Export Correlation: If the "warm" colors (High Export Index) are clustered at the bottom-right, your thesis is confirmed: High reserves stabilize the currency, which leads to better export performance.
